---
title: "Choropleths"
---

## Introduction
This project looked at the rate of stop and frisk reports in Washington DC across both age and gender as well as at national crime rates across the country.[^1] The plots use shapefiles of Washington DC police districts and service areas,[^2], stop and frisk reports,[^3] and national crime rate statistics.[^4] All shapefiles and raw data used are on the associated Github repo. 

[^1]: These plots are based on an in-class case study developed by Dr. Nicole Meyer.
[^2]: The shapefiles were made available by Open Data DC. The police district shapefiles are available [here](http://opendata.dc.gov/datasets/police-districts/) and the police service  area shapefiles are available [here](http://opendata.dc.gov/datasets/police-service-areas/).
[^3]: The stop and frisk reports were made available by the Washington DC Metropolitan Police Department and can be accessed [here](https://mpdc.dc.gov/node/1310236). 
[^4]: National crime statistics were published by the FBI as part of the Uniform Crime Reports and are available [here](https://www.ucrdatatool.gov/).

```{r, message=FALSE, warning = FALSE, include=TRUE, results="hide", echo=TRUE}
# Libraries
library(tidyverse)
library(lubridate)
library(broom)
library(rgdal)
library(gpclib)
library(maptools)
gpclibPermit()
library(maptools)
library(treemap)

# Data
sf_data <- read_csv("sf.csv")
sf_data_2017 <- read_csv("sf2017.csv")

crime_data <- read_csv('CrimeStatebyState.csv')
districts <- readOGR("Police_Districts.shp",layer="Police_Districts")
```

# Data cleaning
  
To clean the data I set age breaks for stop and frisk reports, associated each stop with a location (service district or district), and then summed the number of reports. 
```{r, message=FALSE, warning = FALSE, include=TRUE, results="hide", echo=TRUE}
# bind them together
sf_data <- rbind(sf_data,sf_data_2017) %>% 
   mutate(date = mdy(Report_taken_date_EST),
                  Type=as.factor(`Data Type`),
                  Gender=as.factor(Subject_Sex),
                  Ethnicity=as.factor(Subject_Ethnicity),
                  Race=as.factor(Subject_Race),
                  District=as.factor(`Incident Location District`),
                  PSA=as.factor(`Incident Location PSA`),
                  Year=as.factor(Year)) %>%
  mutate(num_age = ifelse(Age=="Juvenile"|Age=="Unknown",NA,as.numeric(Age))) %>%
  mutate(cat_age = cut(num_age, 
                       breaks=c(17,27,37,47,57,67,77,87),
                       labels=c("18-27","28-37","38-47",
                                                  "48-57","58-67","68-77",
                                                  "78-87"))) %>%
  mutate(cat_age = ifelse(Age=="Juvenile"|Age=="Unknown", Age,
                                            as.character(cat_age))) %>% 
  mutate(Age_binned=as.factor(cat_age)) %>% 
  mutate(Age=fct_relevel(Age,"Juvenile"),
                  Age_binned=fct_relevel(Age_binned,"Juvenile")) %>%
  mutate(Month=month(date,label=T,abbr=F))

remove(sf_data_2017)

# defining districts as "id" variable
names(districts@data)[1] <- "id"

# extracting the lat/long coordinates of district boundary points
# telling the tidy function that the region value can be found in the "id" variable
districts.points <- tidy(districts, region="id")

# combine lat/long coordinates of district boundary points with data about districts
districts.cart <- full_join(districts.points, districts@data, by="id")

# import geospatial data about Police Service Areas
psas <- readOGR("Police_Service_Areas.shp",layer="Police_Service_Areas")

# extracting the lat/long coordinates of PSA boundary points
# telling the tidy function that the region value can be found in the "PSA" variable
psas.points <- tidy(psas, region="PSA")

# combine lat/long coordinates of district boundary points with data about districts
# tell join function that "id" in psas.points corresponds to "PSA" in psas@data
psas.cart <- full_join(psas.points, psas@data, by=c("id" = "PSA"))
```

## Visualizations

```{r plot 1,message=FALSE, warning = FALSE, include=TRUE, results="hide", echo=TRUE}
# A panel of maps showing the number of stop and frisk reports in each police service area faceted by gender
plot_data <- filter(sf_data,!is.na(District), Subject_Sex == 'Male'| Subject_Sex == 'Female')
levels(plot_data$Subject_Sex) <- c('Male', 'Female')

all_cats <- expand(plot_data,PSA,Year)

plot_data <- summarize(group_by(plot_data, PSA,Year,Subject_Sex),reports=n())

plot_data <- full_join(plot_data,all_cats,by=c("PSA","Year"))

plot_data <- mutate(plot_data,reports=replace_na(reports,0))

names(plot_data)[1] <- "NAME"

mapdata <- full_join(plot_data,psas.cart,by="NAME")
levels(mapdata$Subject_Sex) <- c('Male', 'Female')

ggplot() +
  geom_polygon(data=drop_na(mapdata), aes(x=long, y=lat, group=group,fill=reports),color="black") +
  coord_map()+
  labs(title = 'Panel of Stop and Frisk reports per service area faceted by gender')+
  scale_fill_distiller(palette="Spectral", name="Number of Reports")+
  facet_wrap(~Subject_Sex)+
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())

```

```{r plot 2, message=FALSE, warning = FALSE, include=TRUE, results="hide", echo=TRUE}
# A panel of maps showing the number of stop and frisk reports in each police district faceted by age group
plot_data <- filter(sf_data,!is.na(District), !Age_binned == "Unknown")
plot_data$Age_binned <- factor(plot_data$Age_binned)

levels(plot_data$District) <- c("First District","Second District",
                                "Third District","Fourth District",
                                "Fifth District","Sixth District",
                                "Seventh District")

all_cats <- expand(plot_data,District,Year)

plot_data <-summarize(group_by(plot_data, District,Year, Age_binned, .drop=FALSE), reports=n())


plot_data <- full_join(plot_data,all_cats,by=c("District","Year"))

names(plot_data)[1] <- "NAME"

mapdata <- full_join(plot_data,districts.cart,by="NAME")

ggplot() +
  geom_polygon(data=mapdata,
               aes(x=long, y=lat, group=group,fill=reports),color="black") +
  coord_map()+
  theme_void()+
  labs(title = 'Panel of Stop and Frisk reports per police district faceted by age group')+
  scale_fill_distiller(palette="Spectral", name="Number of Reports",na.value="grey")+
  facet_wrap(vars(Age_binned))+
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())
```

```{r plot 3, message=FALSE, warning = FALSE, include=TRUE, results="hide", echo=TRUE}
crime_2013 <- filter(crime_data,Year==2013, State!="United States-Total") %>% 
  rename(region=State) %>% 
  mutate(region=tolower(region))

all_states <- map_data("state")

stateData <- left_join(all_states,crime_2013,by="region")

ggplot()+
  geom_polygon(data=stateData,
               aes(x=long, y=lat, group = group, fill=`Robbery rate`),
               color="grey50")+
  scale_fill_distiller(palette="Oranges",  limits = c(0, 200),direction=1, na.value = 'white')+
  labs(title = 'Choropleth of the Robbery Rate by State in 2013')+
  coord_map()+
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())
```

## Results
The maps show that the stop and frisk rate in DC varies strongly with geographic location, age, and gender. Stops occur most often in eastern DC and to young males. Females are much less likely to be stopped than males in all parts of DC. Similarly, the rate of stops drops off dramatically for people above the age of 40.   

The third visualization, of robberies across the US by state, shows that robberies are geographically heterogenous. Nevada has by far the most robberies per person. Aside from this outlier, states with a large urban population are much more likely to have a high robbery rate per person (e.g. Illinois).


